Differential expression analysis for trisomy 8, deletion 5 and deletion 7
========================================================

> To knit .rmd file, read data files in using "../data"  
> To run chunks in Rstudio, read data files in using "./data"

This code performs differential expression analysis on cleaned RNA-seq RPKM data. In particular, it tests whether there is differential expression between different *deletions* (trisomy 8, del 5 and del 7) using `voom`.


## Load data and required libraries
Load RNA-seq data and the experimental design files:
```{r}
rDes <- read.table("../data/experimental_design_cleaned.txt", 
                           sep = "\t", header = TRUE, row.names = 1) #might need to fix pathname
rDat <- read.table("../data/aml.rnaseq.gaf2.0_rpkm_cleaned.txt",
                          sep = "\t", header = TRUE, row.names = 1) #might need to fix pathname
```

Load required libraries:
```{r warning=FALSE, message=FALSE}
library(lattice) # if you don't already have this loaded ...
library(ggplot2) # we'll make figures with both
library(reshape2) # for the function melt
library(limma)
library(edgeR)
library(car)
library(RColorBrewer)
```

## Data inspection
```{r}
str(rDat, max.level = 0)
rDat[1:4, 1:4]
head(names(rDat))
head(rownames(rDat), n = 10)
tail(rownames(rDat), n = 10)
str(rDes, max.level = 0)
head(rDes)
```

Both the design matrix and the data matrix have already been cleaned, filtered and normalized.

RNA-seq data: there are `r nrow(rDat)` transcripts (rows) for `r length(rDat)` patients (columns). Experimental design: there are `r nrow(rDes)` rows, representing information for each of the patients with RNA-seq data in the AML TCGA data set, and `r length(rDat)` variables.

```

```{r include=FALSE}
# # exploring data via boxplot
# rDatMelt <- melt(rDat, variable.name = "Sample", 
#                  value.name = "RPKM")
# ggplot(rDatMelt, aes(Sample, RPKM)) +
#   geom_boxplot()
```

### Differential expression analysis

I will use `voom` to perform differential expression analysis.

**Deletion**
Which genes are differentially expressed between trisomy 8, deletion 5, deletion 7?

Which genes are differentially expressed between trisomy 8, deletion 5, deletion 7?
```{r}
trisomy8 <- rDes$trisomy_8
table(trisomy8)

del5 <- rDes$del_5
table(del5)

del7 <- rDes$del_7
table(del7)
```

Preparing model matrix
```{r}
#Preparing Model matrices for Voom normalization-------------
modelMat_noInt <- model.matrix(~ trisomy_8*del_5*del_7, rDes)
#The intercept represents a group with all samples without deletion of interest.
#this group changes depending on the deletion, since some samples share multiple deletions.
head(modelMat_noInt)
```

Now find genes differentially expressed between deletion types:
```{r}
dat.voomed <- voom(rDat, modelMat_noInt, plot = TRUE, lib.size = colSums(rDat))
fit_lm <- lmFit(dat.voomed, modelMat_noInt)
fit <- eBayes(fit_lm)
```

Investigating hits (with no interaction)
```{r}
ttfit_tris8 <- topTable(fit, number = Inf, coef = "trisomy_8TRUE", p.value=1e-5)
ttfit_del5 <- topTable(fit, number = Inf, coef = "del_5TRUE", p.value=1e-5)
ttfit_del7 <- topTable(fit, number = Inf, coef = "del_7TRUE", p.value=1e-5)
```
there are:
`r nrow(ttfit_tris8)` hits for trisomy 8, `r nrow(ttfit_del5)` hits for del 5, and `r nrow(ttfit_del7)` hits for del 7.

Is there overlap between the hits?
```{r}
a <- length(intersect(rownames(ttfit_tris8), rownames(ttfit_del5)))
b <- length(intersect(rownames(ttfit_tris8), rownames(ttfit_del7)))
c <- length(intersect(rownames(ttfit_del5), rownames(ttfit_del7)))
sum(a, b, c)
```
there is: `r sum(a, b, c)` overlapping gene, at `intersect(rownames(ttfit_del5), rownames(ttfit_del7)))`

Investigating hits with interaction at 1e-55
```{r}
a <- nrow(ttfit_t8d5 <- topTable(fit, number = Inf, coef = "trisomy_8TRUE:del_5TRUE", p.value=1e-5))
b <- nrow(ttfit_t8d7 <- topTable(fit, number = Inf, coef = "trisomy_8TRUE:del_7TRUE", p.value=1e-5))
c <- nrow(ttfit_d5d7 <- topTable(fit, number = Inf, coef = "del_5TRUE:del_7TRUE", p.value=1e-5))
d <- nrow(ttfit_t8d5d7 <- topTable(fit, number = Inf, coef = "trisomy_8TRUE:del_5TRUE:del_7TRUE", p.value=1e-5))
sum(a, b, c, d)
```
there are `r sum(a, b, c, d)` genes with differential expression influenced by interaction effects (with cutoff of 1e-5).

Investigating hits with interaction at 1e-4
```{r}
a <- nrow(ttfit_t8d5 <- topTable(fit, number = Inf, coef = "trisomy_8TRUE:del_5TRUE", p.value=1e-4))
b <- nrow(ttfit_t8d7 <- topTable(fit, number = Inf, coef = "trisomy_8TRUE:del_7TRUE", p.value=1e-4))
c <- nrow(ttfit_d5d7 <- topTable(fit, number = Inf, coef = "del_5TRUE:del_7TRUE", p.value=1e-4))
d <- nrow(ttfit_t8d5d7 <- topTable(fit, number = Inf, coef = "trisomy_8TRUE:del_5TRUE:del_7TRUE", p.value=1e-4))
sum(a, b, c, d)
```
there is `r sum(a, b, c, d)` gene with differential expression influenced by interaction effects. It is `r rownames(ttfit_t8d7)`

Does it come up in any of the other lists?
```{r}
a <- length(intersect("KCNH6|81033_calculated", rownames(ttfit_tris8)))
b <- length(intersect("KCNH6|81033_calculated", rownames(ttfit_del5)))
c <- length(intersect("KCNH6|81033_calculated", rownames(ttfit_del7)))
sum(a, b, c)
```
it occurs in `r sum(a, b, c)` of the other lists. 

*Exploring hit genes from trisomy_8.*
topTable for Trisomy 8

```{r}
(ttfit_tris8)
```

Plotsmear of trisomy_8 hits.

```{r tidy=FALSE}
# Create a DGEList object
voomTris8genes <- rownames(ttfit_tris8)
trisomy_8 <- rDes$trisomy_8
dgeGlmT8 <- DGEList(counts = rDat, group = as.numeric(trisomy_8))
plotSmear(dgeGlmT8, de.tags = voomTris8genes, ylab = "logFC", xlab = "AverageRPKM", main = "RPKM of genes differentially expressed in trisomy_8 samples")
abline(h = c(-1, 1), col = "blue")
```

Creating a boxplot with the `r length(voomTris8genes)` genes of interest (FDR 1e-5) for trisomy 8

```{r tidy=FALSE}
#subsetting and reforming transcripts of interest from main RPKM matrix
tris8Dat <- rDat[voomTris8genes,]
tris8Dat$Transcript <- rownames(tris8Dat) #creating a transcript group
tris8Dat <- melt(tris8Dat, id.vars = "Transcript", 
                   variable.name = "TCGA_patient_id",
                   value.name = "RPKM")

#cleaup molten RPKM data
tris8Dat$TCGA_patient_id <- gsub("X", "", tris8Dat$TCGA_patient_id)
tris8Dat$Transcript <- gsub("[|].*$", "", tris8Dat$Transcript)

#preparing a small design matrix and merging
miniDes <- rDes[,c("trisomy_8", "del_5" , "del_7")]
miniDes$"TCGA_patient_id" <- rownames(rDes)
tris8Dat <- merge(tris8Dat, miniDes, by = "TCGA_patient_id") #merging

#plotting
ggplot(tris8Dat, aes(Transcript, log2(RPKM), colour = trisomy_8)) +
  geom_boxplot() +
  facet_wrap(~ Transcript, scales = "free") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

*Exploring hit genes from del_5.*
TopTable for del_5

```{r}
(ttfit_del5)
```

Plotsmear of del_5 hits

```{r echo = FALSE}
# Create a DGEList object
voomDel5genes <- rownames(ttfit_del5)
del_5 <- rDes$del_5
dgeGlmD5 <- DGEList(counts = rDat, group = as.numeric(del_5))
plotSmear(dgeGlmD5, de.tags = voomDel5genes, ylab = "logFC", xlab = "AverageRPKM", main = "RPKM of genes differentially expressed in del_5 samples")
abline(h = c(-1, 1), col = "blue")
```

Creating a boxplot with the `r length(voomDel5genes)` genes of interest (FDR 1e-5) for del 5

```{r echo = FALSE}
del5Dat <- rDat[voomDel5genes,] #subsetting transcripts of interest from rDat
del5Dat$Transcript <- rownames(del5Dat) #creating a transcript group
del5Dat <- melt(del5Dat, id.vars = "Transcript", 
                   variable.name = "TCGA_patient_id",
                   value.name = "RPKM")

del5Dat$TCGA_patient_id <- gsub("X", "", del5Dat$TCGA_patient_id) #removing the "X" preceding patient ID names
del5Dat$Transcript <- gsub("[|].*$", "", del5Dat$Transcript) #shortening transcript names

#miniDes already made while exploring trisomy 8.
del5Dat <- merge(del5Dat, miniDes, by = "TCGA_patient_id") #merging

ggplot(del5Dat, aes(Transcript, log2(RPKM), colour = del_5)) +
  geom_boxplot()  +
  facet_wrap(~ Transcript, scales = "free") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

Creating a boxplot with the genes of interest (FDR 1e-6) for del 5
```{r echo = FALSE}
voomDel5genes <- rownames(ttfit_del5[(which(ttfit_del5$adj.P.Val < 1e-6)),])
del5Dat <- rDat[voomDel5genes,] #subsetting transcripts of interest from rDat
del5Dat$Transcript <- rownames(del5Dat) #creating a transcript group
del5Dat <- melt(del5Dat, id.vars = "Transcript", 
                   variable.name = "TCGA_patient_id",
                   value.name = "RPKM")

del5Dat$TCGA_patient_id <- gsub("X", "", del5Dat$TCGA_patient_id) #removing the "X" preceding patient ID names
del5Dat$Transcript <- gsub("[|].*$", "", del5Dat$Transcript) #shortening transcript names

#miniDes already made while exploring trisomy 8.
del5Dat <- merge(del5Dat, miniDes, by = "TCGA_patient_id") #merging

ggplot(del5Dat, aes(Transcript, log2(RPKM), colour = del_5)) +
  geom_boxplot()  +
  facet_wrap(~ Transcript, scales = "free") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

testing expression level cutoff: Removing genes with max RPKM of less than 8
```{r echo = FALSE}
del5Dat <- rDat[voomDel5genes,] #subsetting transcripts of interest from rDat
del5Dat <- del5Dat[(apply(del5Dat, 1, max) > 8),] #or something like this
del5Dat$Transcript <- rownames(del5Dat) #creating a transcript group
del5Dat <- melt(del5Dat, id.vars = "Transcript", 
                   variable.name = "TCGA_patient_id",
                   value.name = "RPKM")

del5Dat$TCGA_patient_id <- gsub("X", "", del5Dat$TCGA_patient_id) #removing the "X" preceding patient ID names
del5Dat$Transcript <- gsub("[|].*$", "", del5Dat$Transcript) #shortening transcript names

#miniDes already made while exploring trisomy 8.
del5Dat <- merge(del5Dat, miniDes, by = "TCGA_patient_id") #merging

ggplot(del5Dat, aes(Transcript, log2(RPKM), colour = del_5)) +
  geom_boxplot()  +
  facet_wrap(~ Transcript, scales = "free") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

*Exploring hit genes from del_7.*
TopTable for genes from del_7
```{r}
head(ttfit_del7)
nrow(ttfit_del7)
tail(ttfit_del7)
```

Plotsmear of del_7 hits
```{r echo = FALSE}
# Create a DGEList object
voomDel7genes <- rownames(ttfit_del7)
del_7 <- rDes$del_7
dgeGlmD7 <- DGEList(counts = rDat, group = as.numeric(del_7))
plotSmear(dgeGlmD7, de.tags = voomDel7genes, ylab = "logFC", xlab = "AverageRPKM", main = "RPKM of genes differentially expressed in del_7 samples")
abline(h = c(-1, 1), col = "blue")
```

Creating a boxplot with genes of interest (FDR 1e-5) for del 7. Since there are `r length(voomDel7genes)` genes I will plot a few genes at a time:
```{r echo = FALSE}
del7Dat <- rDat[voomDel7genes[1:16],] #subsetting transcripts of interest from rDat
del7Dat$Transcript <- rownames(del7Dat) #creating a transcript group
del7Dat <- melt(del7Dat, id.vars = "Transcript", 
                   variable.name = "TCGA_patient_id",
                   value.name = "RPKM")

del7Dat$TCGA_patient_id <- gsub("X", "", del7Dat$TCGA_patient_id) #removing the "X" preceding patient ID names
del7Dat$Transcript <- gsub("[|].*$", "", del7Dat$Transcript) #shortening transcript names

#miniDes already made while exploring trisomy 8.
del7Dat <- merge(del7Dat, miniDes, by = "TCGA_patient_id") #merging

ggplot(del7Dat, aes(Transcript, log2(RPKM), colour = del_7)) +
  geom_boxplot()  +
  facet_wrap(~ Transcript, scales = "free") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

```{r echo = FALSE}
del7Dat <- rDat[voomDel7genes[17:32],] #subsetting transcripts of interest from rDat
del7Dat$Transcript <- rownames(del7Dat) #creating a transcript group
del7Dat <- melt(del7Dat, id.vars = "Transcript", 
                   variable.name = "TCGA_patient_id",
                   value.name = "RPKM")

del7Dat$TCGA_patient_id <- gsub("X", "", del7Dat$TCGA_patient_id) #removing the "X" preceding patient ID names
del7Dat$Transcript <- gsub("[|].*$", "", del7Dat$Transcript) #shortening transcript names

#miniDes already made while exploring trisomy 8.
del7Dat <- merge(del7Dat, miniDes, by = "TCGA_patient_id") #merging

ggplot(del7Dat, aes(Transcript, log2(RPKM), colour = del_7)) +
  geom_boxplot()  +
  facet_wrap(~ Transcript, scales = "free") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

```{r echo = FALSE}
del7Dat <- rDat[voomDel7genes[33:48],] #subsetting transcripts of interest from rDat
del7Dat$Transcript <- rownames(del7Dat) #creating a transcript group
del7Dat <- melt(del7Dat, id.vars = "Transcript", 
                   variable.name = "TCGA_patient_id",
                   value.name = "RPKM")

del7Dat$TCGA_patient_id <- gsub("X", "", del7Dat$TCGA_patient_id) #removing the "X" preceding patient ID names
del7Dat$Transcript <- gsub("[|].*$", "", del7Dat$Transcript) #shortening transcript names

#miniDes already made while exploring trisomy 8.
del7Dat <- merge(del7Dat, miniDes, by = "TCGA_patient_id") #merging

ggplot(del7Dat, aes(Transcript, log2(RPKM), colour = del_7)) +
  geom_boxplot()  +
  facet_wrap(~ Transcript, scales = "free") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

```{r echo = FALSE}
del7Dat <- rDat[voomDel7genes[49:61],] #subsetting transcripts of interest from rDat
del7Dat$Transcript <- rownames(del7Dat) #creating a transcript group
del7Dat <- melt(del7Dat, id.vars = "Transcript", 
                   variable.name = "TCGA_patient_id",
                   value.name = "RPKM")

del7Dat$TCGA_patient_id <- gsub("X", "", del7Dat$TCGA_patient_id) #removing the "X" preceding patient ID names
del7Dat$Transcript <- gsub("[|].*$", "", del7Dat$Transcript) #shortening transcript names

#miniDes already made while exploring trisomy 8.
del7Dat <- merge(del7Dat, miniDes, by = "TCGA_patient_id") #merging

ggplot(del7Dat, aes(Transcript, log2(RPKM), colour = del_7)) +
  geom_boxplot()  +
  facet_wrap(~ Transcript, scales = "free") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```