SVM Predictions
========================================================

### Prep Work

**Load RNA Seq data and design, then merge. I will use the read counts data for this analysis.**
```{r}
rna_seq_dat <-read.table("../../data/aml.rnaseq.gaf2.0_rpkm_cleaned.txt", sep = "\t", 
                            header = TRUE, check.names = FALSE)

exp_dat <- read.table("../../data/experimental_design_cleaned.txt", 
                           sep = "\t", header = TRUE, row.names = 1)

joined_dat <- cbind(exp_dat,t(rna_seq_dat))
```

**Load libraries.**
```{r}
library(plyr)
library(kernlab)
library(taRifx)
library(cvTools)
library(VennDiagram)
library(limma)
library(edgeR)
```

**Helper functions.**
```{r}
# Function to select features using linear models
# input.dat: training data set
# input.labels: true outcomes for training data
fs.lm <- function(input.dat, input.labels) {
  norm.factor <- calcNormFactors(input.dat)
  design <- model.matrix(~input.labels)[,1:2]
  colnames(design) <- c("Intercept", "Label")
  dat.voomed <- voom(input.dat, design, lib.size = colSums(input.dat) * norm.factor)
  fit <- lmFit(dat.voomed, design)
  ebFit <- eBayes(fit)
  hits <- topTable(ebFit, n = Inf, coef = "Label")
  train.features <- hits$ID[1:25]
  return(train.features)
}
```

Part 1: Predicting Cytogenetic Risk
-------------------------

### Part 1a: Data prep

**First, let's reorder the factor levels for Cytogenetic_risk so they make sense.**
```{r}
joined_dat$Cytogenetic_risk <- factor(joined_dat$Cytogenetic_risk, levels(joined_dat$Cytogenetic_risk)[c(4,2,1,3)])
```

**Remove the rows with N.D.**
```{r}
dat_filt <- subset(joined_dat,joined_dat$Cytogenetic_risk != "N.D.")
```

**Now, for the SVM, let's make a binary prognosis columns (TRUE for poor).**
```{r}
dat_svm <- dat_filt
dat_svm$prognosis <- mapvalues(dat_svm$Cytogenetic_risk, c("Good","Intermediate","Poor"), c(FALSE,FALSE,TRUE), warn_missing = TRUE)
```

**Make prelimary dataset with gene columns + prognosis + cyt risk.**
```{r}
prelim <- dat_svm[,grepl(".*calculated",colnames(dat_svm))]
prelim <- dat_svm[,c(colnames(prelim),"prognosis","Cytogenetic_risk")]
```

### Part 1b: Predict using just most correlated gene features

**Prepare folds and confusion matrix for cross validation.**
```{r}
set.seed(540)
folds <- cvFolds(nrow(prelim), K = 5)
conf_matrix <- matrix(0,nrow=2,ncol=2,dimnames=list(c("truePoor","trueNotPoor"),c("predPoor","predNotPoor")))
```

**Prepare list for top genes.**
```{r}
top_genes_per_fold <- list()
```

**Perform cross validation.**
```{r}
for (f in 1:5) {

  # Divide preliminary dataset into train and test sets.
  prelim_train <- prelim[folds$subsets[folds$which!=f,],]
  prelim_test <- prelim[folds$subsets[folds$which==f],]
  prelim_test_labels <- prelim_test$prognosis
  
  # *** Narrow the feature set by finding the genes most correlated with risk in the test set ***
  
  # Get just gene columns
  gene_cols <- prelim_train[,grepl(".*calculated",colnames(prelim_train))]
  
  # Get correlations
  gene_corrs <- as.data.frame(t(cor(as.numeric(prelim_train$Cytogenetic_risk),gene_cols,method="spearman")))
  
  # Sort
  gene_corrs_sort <- sort(abs(gene_corrs), f= ~ -V1, drop=FALSE)
  
  # Grab top 25
  gene_corrs_top <- gene_corrs_sort[1:25,,drop=FALSE]
  top_genes_per_fold <- c(top_genes_per_fold,rownames(gene_corrs_top))
  
  # *** Do SVM ***
  
  # Prepare training and test sets
  dat_svm_gene_train <- prelim_train[,c(rownames(gene_corrs_top),"prognosis")]
  dat_svm_gene_test <- prelim_test[,c(rownames(gene_corrs_top),"prognosis")]
  
  # Build
  fit_svm <- ksvm(prognosis~.,dat_svm_gene_train)
  
  # Predict
  pred_svm <- predict(fit_svm,newdata=dat_svm_gene_test,type="response")
  
  # Process
  results <- table(prelim_test_labels,pred_svm)
  
  conf_matrix[1,1] <- conf_matrix[1,1] + results[1,1]
  conf_matrix[1,2] <- conf_matrix[1,2] + results[1,2]
  conf_matrix[2,1] <- conf_matrix[2,1] + results[2,1]
  conf_matrix[2,2] <- conf_matrix[2,2] + results[2,2]
}
```

**Process CV results.**
```{r}
sens_svm_cor <- conf_matrix[1,1]/sum(conf_matrix[1,])
spec_svm_cor <- conf_matrix[2,2]/sum(conf_matrix[2,])
sink("results_risk.txt")
cat("Risk, corrs:")
cat("\n")
cat(sens_svm_cor)
cat("\n")
cat(spec_svm_cor)
cat("\n")
cat("\n")
```

**Explore the genes selected in each fold.**
```{r}
fold1 <- top_genes_per_fold[1:25];
fold2 <- top_genes_per_fold[26:50];
fold3 <- top_genes_per_fold[51:75];
fold4 <- top_genes_per_fold[76:100];
fold5 <- top_genes_per_fold[101:125];

folds <- list(Fold1=fold1,Fold2=fold2,Fold3=fold3,Fold4=fold4,Fold5=fold5)
plot.new()
venn_plot <- venn.diagram(folds, filename = NULL,force.unique=TRUE,ext.text=FALSE,margin=0.1)
grid.draw(venn_plot)

intersection_cor <- unlist(intersect(fold1,intersect(fold2,intersect(fold3,intersect(fold4,fold5)))))
cat(intersection_cor)
cat("\n\n")
sink(file=NULL)
```

### Part 1c: Predict using differentially expressed features (according to linear model analysis)

**Prepare folds and confusion matrix for cross validation.**
```{r}
set.seed(540)
folds <- cvFolds(nrow(prelim), K = 5)
conf_matrix <- matrix(0,nrow=2,ncol=2,dimnames=list(c("truePoor","trueNotPoor"),c("predPoor","predNotPoor")))
```

**Prepare list for top genes.**
```{r}
selections_per_fold <- list()
```

**Perform cross validation.**
```{r}
for (f in 1:5) {

  # Divide preliminary dataset into train and test sets.
  prelim_train <- prelim[folds$subsets[folds$which!=f,],]
  prelim_test <- prelim[folds$subsets[folds$which==f],]
  prelim_test_labels <- prelim_test$prognosis
  
  # *** Narrow the feature set by finding the top differentially expressed genes ***
  
  # Get just gene columns
  gene_cols <- prelim_train[,grepl(".*calculated",colnames(prelim_train))]
  
  # Do lm selection
  selections <- fs.lm(t(gene_cols),prelim_train$prognosis)
  selections_per_fold <- c(selections_per_fold,selections)
  
  # *** Do SVM ***
  
  # Prepare training and test sets
  dat_svm_gene_train <- prelim_train[,c(selections,"prognosis")]
  dat_svm_gene_test <- prelim_test[,c(selections,"prognosis")]
  
  # Build
  fit_svm <- ksvm(prognosis~.,dat_svm_gene_train)
  
  # Predict
  pred_svm <- predict(fit_svm,newdata=dat_svm_gene_test,type="response")
  
  # Process
  results <- table(prelim_test_labels,pred_svm)
  
  conf_matrix[1,1] <- conf_matrix[1,1] + results[1,1]
  conf_matrix[1,2] <- conf_matrix[1,2] + results[1,2]
  conf_matrix[2,1] <- conf_matrix[2,1] + results[2,1]
  conf_matrix[2,2] <- conf_matrix[2,2] + results[2,2] 
}
```

**Process CV results.**
```{r}
sens_svm_lm <- conf_matrix[1,1]/sum(conf_matrix[1,])
spec_svm_lm <- conf_matrix[2,2]/sum(conf_matrix[2,])
sink("results_risk.txt",append=TRUE)
cat("Risk, lm:")
cat("\n")
cat(sens_svm_lm)
cat("\n")
cat(spec_svm_lm)
cat("\n")
cat("\n")
```

**Explore the genes selected in each fold.**
```{r}
fold1 <- selections_per_fold[1:25];
fold2 <- selections_per_fold[26:50];
fold3 <- selections_per_fold[51:75];
fold4 <- selections_per_fold[76:100];
fold5 <- selections_per_fold[101:125];

folds <- list(Fold1=fold1,Fold2=fold2,Fold3=fold3,Fold4=fold4,Fold5=fold5)
plot.new()
venn_plot <- venn.diagram(folds, filename = NULL,force.unique=TRUE,ext.text=FALSE,margin=0.1)
grid.draw(venn_plot)

intersection_lm <- unlist(intersect(fold1,intersect(fold2,intersect(fold3,intersect(fold4,fold5)))))
cat(intersection_lm)
cat("\n\n")
cat(intersect(intersection_cor,intersection_lm))
cat("\n\n")
sink(file=NULL)
```

Part 2: Predicting Cytogenetic Features (trisomy_8, del_5, and del_7)
-------------------------

### Part 2a: Data prep

**Copy the original data and factorize the features columns (trisomy_8/del_5/del_7).**
```{r}
dat_svm <- joined_dat
dat_svm$trisomy_8 <- as.factor(dat_svm$trisomy_8)
dat_svm$del_5 <- as.factor(dat_svm$del_5)
dat_svm$del_7 <- as.factor(dat_svm$del_7)
```

**Make prelimary dataset with gene columns + prognosis + cyt risk.**
```{r}
prelim <- dat_svm[,grepl(".*calculated",colnames(dat_svm))]
#prelim$prognosis <- dat_svm$trisomy_8
#prelim$prognosis <- dat_svm$del_5
prelim$prognosis <- dat_svm$del_7
```

### Part 2b: Predict using just most correlated gene features

**Prepare folds and confusion matrix for cross validation.**
```{r}
set.seed(540)
folds <- cvFolds(nrow(prelim), K = 5)
conf_matrix <- matrix(0,nrow=2,ncol=2,dimnames=list(c("truePoor","trueNotPoor"),c("predPoor","predNotPoor")))
```

**Prepare list for top genes.**
```{r}
top_genes_per_fold <- list()
```

**Perform cross validation.**
```{r}
for (f in 1:5) {

  # Divide preliminary dataset into train and test sets.
  prelim_train <- prelim[folds$subsets[folds$which!=f,],]
  prelim_test <- prelim[folds$subsets[folds$which==f],]
  prelim_test_labels <- prelim_test$prognosis
  
  # *** Narrow the feature set by finding the genes most correlated with risk in the test set ***
  
  # Get just gene columns
  gene_cols <- prelim_train[,grepl(".*calculated",colnames(prelim_train))]
  
  # Get correlations
  gene_corrs <- as.data.frame(t(cor(as.numeric(prelim_train$prognosis),gene_cols,method="spearman")))
  
  # Sort
  gene_corrs_sort <- sort(gene_corrs, f= ~ -V1, drop=FALSE)
  
  # Grab top 25
  gene_corrs_top <- gene_corrs_sort[1:25,,drop=FALSE]
  top_genes_per_fold <- c(top_genes_per_fold,rownames(gene_corrs_top))
  
  # *** Do SVM ***
  
  # Prepare training and test sets
  dat_svm_gene_train <- prelim_train[,c(rownames(gene_corrs_top),"prognosis")]
  dat_svm_gene_test <- prelim_test[,c(rownames(gene_corrs_top),"prognosis")]
  
  # Build
  fit_svm <- ksvm(prognosis~.,dat_svm_gene_train)
  
  # Predict
  pred_svm <- predict(fit_svm,newdata=dat_svm_gene_test,type="response")
  
  # Process
  results <- table(prelim_test_labels,pred_svm)
  
  conf_matrix[1,1] <- conf_matrix[1,1] + results[2,2]
  conf_matrix[1,2] <- conf_matrix[1,2] + results[2,1]
  conf_matrix[2,1] <- conf_matrix[2,1] + results[1,2]
  conf_matrix[2,2] <- conf_matrix[2,2] + results[1,1]
}
```

**Process CV results.**
```{r}
sens_svm_cor <- conf_matrix[1,1]/sum(conf_matrix[1,])
spec_svm_cor <- conf_matrix[2,2]/sum(conf_matrix[2,])
#sink("results_t8.txt")
#sink("results_d5.txt")
sink("results_d7.txt")
#cat("Trisomy8, corrs:")
#cat("Del5, corrs:")
cat("Del7, corrs:")
cat("\n")
cat(sens_svm_cor)
cat("\n")
cat(spec_svm_cor)
cat("\n")
cat("\n")
```

**Explore the genes selected in each fold.**
```{r}
fold1 <- top_genes_per_fold[1:25];
fold2 <- top_genes_per_fold[26:50];
fold3 <- top_genes_per_fold[51:75];
fold4 <- top_genes_per_fold[76:100];
fold5 <- top_genes_per_fold[101:125];

folds <- list(Fold1=fold1,Fold2=fold2,Fold3=fold3,Fold4=fold4,Fold5=fold5)
plot.new()
venn_plot <- venn.diagram(folds, filename = NULL,force.unique=TRUE,ext.text=FALSE,margin=0.1)
grid.draw(venn_plot)

intersection_cor <- unlist(intersect(fold1,intersect(fold2,intersect(fold3,intersect(fold4,fold5)))))
cat(intersection_cor)
cat("\n\n")
sink(file=NULL)
```

### Part 2c: Predict using differentially expressed features (according to linear model analysis)

**Prepare folds and confusion matrix for cross validation.**
```{r}
set.seed(540)
folds <- cvFolds(nrow(prelim), K = 5)
conf_matrix <- matrix(0,nrow=2,ncol=2,dimnames=list(c("truePoor","trueNotPoor"),c("predPoor","predNotPoor")))
```

**Prepare list for top genes.**
```{r}
selections_per_fold <- list()
```

**Perform cross validation.**
```{r}
for (f in 1:5) {

  # Divide preliminary dataset into train and test sets.
  prelim_train <- prelim[folds$subsets[folds$which!=f,],]
  prelim_test <- prelim[folds$subsets[folds$which==f],]
  prelim_test_labels <- prelim_test$prognosis
  
  # *** Narrow the feature set by finding the top differentially expressed genes ***
  
  # Get just gene columns
  gene_cols <- prelim_train[,grepl(".*calculated",colnames(prelim_train))]
  
  # Do lm selection
  selections <- fs.lm(t(gene_cols),prelim_train$prognosis)
  selections_per_fold <- c(selections_per_fold,selections)
  
  # *** Do SVM ***
  
  # Prepare training and test sets
  dat_svm_gene_train <- prelim_train[,c(selections,"prognosis")]
  dat_svm_gene_test <- prelim_test[,c(selections,"prognosis")]
  
  # Build
  fit_svm <- ksvm(prognosis~.,dat_svm_gene_train)
  
  # Predict
  pred_svm <- predict(fit_svm,newdata=dat_svm_gene_test,type="response")
  
  # Process
  results <- table(prelim_test_labels,pred_svm)
  
  conf_matrix[1,1] <- conf_matrix[1,1] + results[2,2]
  conf_matrix[1,2] <- conf_matrix[1,2] + results[2,1]
  conf_matrix[2,1] <- conf_matrix[2,1] + results[1,2]
  conf_matrix[2,2] <- conf_matrix[2,2] + results[1,1] 
}
```

**Process CV results.**
```{r}
sens_svm_lm <- conf_matrix[1,1]/sum(conf_matrix[1,])
spec_svm_lm <- conf_matrix[2,2]/sum(conf_matrix[2,])
#sink("results_t8.txt",append=TRUE)
#sink("results_d5.txt",append=TRUE)
sink("results_d7.txt",append=TRUE)
#cat("Trisomy8, lm:")
#cat("Del5, lm:")
cat("Del7, lm:")
cat("\n")
cat(sens_svm_lm)
cat("\n")
cat(spec_svm_lm)
cat("\n")
cat("\n")
```

**Explore the genes selected in each fold.**
```{r}
fold1 <- selections_per_fold[1:25];
fold2 <- selections_per_fold[26:50];
fold3 <- selections_per_fold[51:75];
fold4 <- selections_per_fold[76:100];
fold5 <- selections_per_fold[101:125];

folds <- list(Fold1=fold1,Fold2=fold2,Fold3=fold3,Fold4=fold4,Fold5=fold5)
plot.new()
venn_plot <- venn.diagram(folds, filename = NULL,force.unique=TRUE,ext.text=FALSE,margin=0.1)
grid.draw(venn_plot)

intersection_lm <- unlist(intersect(fold1,intersect(fold2,intersect(fold3,intersect(fold4,fold5)))))
cat(intersection_lm)
cat("\n\n")
cat(intersect(intersection_cor,intersection_lm))
cat("\n\n")
sink(file=NULL)
```