PCA Analysis
================

```{r}

rna_seq_dat_rpkm <-read.table("../../data/aml.rnaseq.gaf2.0_rpkm_cleaned.txt", sep = "\t", header = TRUE, check.names = FALSE)
rna_seq_dat_read <-read.table("../../data/aml.rnaseq.gaf2.0_read_count_cleaned.txt", sep = "\t", header = TRUE, check.names = FALSE)


exp_dat <- read.table("../../data/experimental_design_cleaned.txt", 
                           sep = "\t", header = TRUE, row.names = 1)

joined_dat_rpkm <- cbind(exp_dat,t(rna_seq_dat_rpkm))
joined_dat_read <- cbind(exp_dat,t(rna_seq_dat_read))

```

**Load libraries.**
```{r}
library(plyr)
library(kernlab)
library(taRifx)
library(cvTools)
library(grid)
library(VennDiagram)
```
**Now, let's remove the columns with NA values - we won't use them at this time.**
```{r}
dat.complete.rpkm <- joined_dat_rpkm

dat.complete.read <- joined_dat_read
```


```
**Remove the rows with N.D.**
```{r}
dat.filt.rpkm <- subset(dat.complete.rpkm,dat.complete.rpkm$del_5 != "N.D.")
dat.filt.read <- subset(dat.complete.read,dat.complete.read$del_5 != "N.D.")
```

**let's make a binary prognosis columns (TRUE for poor).**
```{r}
dat.in.read <- dat.filt.read
dat.in.read$prognosis <- mapvalues(dat.in.read$del_5, c("TRUE","FALSE"), c(1,0), warn_missing = TRUE)
dat.in.rpkm <- dat.filt.rpkm
dat.in.rpkm$prognosis <- mapvalues(dat.in.rpkm$del_5, c("TRUE","FALSE"), c(1,0), warn_missing = TRUE)
```

**Make prelimary dataset with gene columns + prognosis + cyt risk.**
```{r}
prelim.in.read <- dat.in.read[,grepl("_*calculated",colnames(dat.in.read))]
gene.name.read <- colnames(prelim.in.read)
prelim.in.read <- dat.in.read[,c(colnames(prelim.in.read),"prognosis","del_5")]

prelim.in.rpkm <- dat.in.rpkm[,grepl("_*calculated",colnames(dat.in.rpkm))]
gene.name.rpkm <- colnames(prelim.in.rpkm)
prelim.in.rpkm <- dat.in.rpkm[,c(colnames(prelim.in.rpkm),"prognosis","del_5")]
```

**kernel PCA**
```{r}
library(geigen)
kernel_pca <- function(x.tr, x.ts, label.tr, d, flag){
  #x.tr and x.ts are feature*sample
  # flag =TRUE is for supervised kernel PCA and flag= False is for kernel pca
  # "d" is the dimension of the data in the new feature space
  n <- dim(x.tr)[2]
  I <- diag(n)
  e <- rep(1, n)
  s <- 1/n
  H <- I - s*tcrossprod(e)
  if (flag==TRUE){
    L <- label.tr %*% t(label.tr)
    cat("Flag is true\n")
    }else{
    cat("Flag is false\n")
    L <- I    
  }
    
  k <- t(x.tr) %*% x.tr
  Q <- k %*% H %*% L %*% H %*% k
  Q.svd <- svd(Q, nu=d, nv=d)
  U <- Q.svd$u[,1:d]
  V <- Q.svd$v[,1:d]
  D <- Q.svd$d[1:d]
  cat("Eigen values =", D , "\n")
  beta <- geigen(Q, k)
  mapped.tr <- beta$vectors %*% k
  mapped.ts <- beta$vectors %*% t(x.tr) %*% x.ts
  
  list(tr.dat=mapped.tr[1:d,], ts.dat=mapped.ts[1:d,])
}

```

**run svm**
```{r}
run_svm <- function(x.tr, x.ts){
  conf_matrix <- matrix(0,nrow=2,ncol=2,dimnames=list(c("truePoor","trueNotPoor"),c("predPoor","predNotPoor")))
  fit_svm <- ksvm(prognosis~.,x.tr)
  pred_svm <- predict(fit_svm,newdata=x.ts,type="response")

  results <- table(x.ts$prognosis,pred_svm)
  print(results)
  conf_matrix[1,1] <- conf_matrix[1,1] + results[1,1]
  conf_matrix[1,2] <- conf_matrix[1,2] + results[1,2]
  conf_matrix[2,1] <- conf_matrix[2,1] + results[2,1]
  conf_matrix[2,2] <- conf_matrix[2,2] + results[2,2]
  
  sens_svm <- conf_matrix[1,1]/sum(conf_matrix[1,])
  spec_svm <- conf_matrix[2,2]/sum(conf_matrix[2,])
  cat("Sensitvity = ", sens_svm, "\n")
  cat("Specificity = ", spec_svm, "\n")
  
}

```

**Perform PCA-SVM by cross validation.**
```{r}
cross_validate <- function(prelim_in, d){ 
  k<- 5
  set.seed(540)
  folds <- cvFolds(nrow(prelim.in.read), K = k)
  conf_matrix_ker <- matrix(0,nrow=2,ncol=2,dimnames=list(c("truePoor","trueNotPoor"),c("predPoor","predNotPoor")))
  conf_matrix_sup <- matrix(0,nrow=2,ncol=2,dimnames=list(c("truePoor","trueNotPoor"),c("predPoor","predNotPoor")))
  conf_matrix_basic <- matrix(0,nrow=2,ncol=2,dimnames=list(c("truePoor","trueNotPoor"),c("predPoor","predNotPoor")))  
  for (f in 1:k) {
    
    x.ts <- t(data.matrix(prelim.in[folds$subsets[folds$which==f,],grepl(".*calculated",colnames(prelim.in))]))
    ytest <- factor(prelim.in[folds$subsets[folds$which==f,],c("prognosis")])
    ytest <- as.numeric(levels(ytest))[ytest]
    x.tr <- t(data.matrix(prelim.in[folds$subsets[folds$which!=f,],grepl(".*calculated",colnames(prelim.in))]))
    ytrain <- factor(prelim.in[folds$subsets[folds$which!=f,],c("prognosis")])
    ytrain <- as.numeric(levels(ytrain))[ytrain]
    cat(ytest,"\n")
    cat(nrow(which(ytest=="1")))
    cat(nrow(which(ytest=="0")))
    
    #kernel PCA
    res <- kernel_pca((x.tr), (x.ts),  ytrain, d, FALSE)
    
    xtrain <- t(res$tr.dat)
    xtest <- t(res$ts.dat)
    
    svp <- ksvm(xtrain,factor(ytrain),type="C-svc",kernel="rbfdot",C=100,scaled=c())
    ypred = predict(svp,xtest)
    results.ker <- table(factor(ytest),ypred)

    
    conf_matrix_ker[1,1] <- conf_matrix_ker[1,1] + results.ker[1,1]
    conf_matrix_ker[1,2] <- conf_matrix_ker[1,2] + results.ker[1,2]
    conf_matrix_ker[2,1] <- conf_matrix_ker[2,1] + results.ker[2,1]
    conf_matrix_ker[2,2] <- conf_matrix_ker[2,2] + results.ker[2,2] 
   
    ###############
    #supervised PCA
    cat("supervised PCA\n")
    sup.res <- kernel_pca((x.tr), (x.ts), ytrain, d, TRUE)
    
    xtrain <- t(res$tr.dat)
    
    xtest <- t(res$ts.dat)
    
    svp <- ksvm(xtrain,factor(ytrain),type="C-svc",kernel="rbfdot",C=100,scaled=c())
    ypred = predict(svp,xtest)
    results.sup <- table(factor(ytest),ypred)
    
    conf_matrix_sup[1,1] <- conf_matrix_sup[1,1] + results.sup[1,1]
    conf_matrix_sup[1,2] <- conf_matrix_sup[1,2] + results.sup[1,2]
    conf_matrix_sup[2,1] <- conf_matrix_sup[2,1] + results.sup[2,1]
    conf_matrix_sup[2,2] <- conf_matrix_sup[2,2] + results.sup[2,2] 
    ##############
    cat("basic PCA\n")
    #basic PCA
    princomp <- prcomp(prelim.in[,grepl(".*calculated",colnames(prelim.in))], center = T, scale = F)

    xtrain <- princomp$x[folds$subsets[folds$which!=f,], 1:d]
    xtest <- princomp$x[folds$subsets[folds$which==f,], 1:d]

    svp <- ksvm(xtrain,factor(ytrain),type="C-svc",kernel="rbfdot",C=100,scaled=c())
    ypred = predict(svp,xtest)
    results.basic <- table(factor(ytest),ypred)
    
    conf_matrix_basic[1,1] <- conf_matrix_basic[1,1] + results.basic[1,1]
    conf_matrix_basic[1,2] <- conf_matrix_basic[1,2] + results.basic[1,2]
    conf_matrix_basic[2,1] <- conf_matrix_basic[2,1] + results.basic[2,1]
    conf_matrix_basic[2,2] <- conf_matrix_basic[2,2] + results.basic[2,2] 
  }
  list (basic=conf_matrix_basic, ker=conf_matrix_ker, sup=conf_matrix_sup)
}
```
```{r}
prepareData  <- function(g, prDat, prDes ){
  pDat <- data.frame()
  for (i in 1:length(g)){
    pDat1 <- data.frame(prDes, gExp = as.vector(as.matrix(prDat[g[i], ])), gene = g[i])
    pDat <- rbind(pDat, pDat1)
  } 
  pDat
}

```



**RPKM data analysis**
```{r}

prelim.in <- prelim.in.rpkm
gene.name <- gene.name.rpkm

rpkm.results <- cross_validate(prelim.in, 20)
#process CV results
conf_matrix_basic <- rpkm.results$basic
conf_matrix_ker <- rpkm.results$ker
conf_matrix_sup <- rpkm.results$sup

(sens_svm_basic <- conf_matrix_basic[1,1]/sum(conf_matrix_basic[1,]))
(spec_svm_basic <- conf_matrix_basic[2,2]/sum(conf_matrix_basic[2,]))


(sens_svm_ker <- conf_matrix_ker[1,1]/sum(conf_matrix_ker[1,]))
(spec_svm_ker <- conf_matrix_ker[2,2]/sum(conf_matrix_ker[2,]))


(sens_svm_sup <- conf_matrix_sup[1,1]/sum(conf_matrix_sup[1,]))
(spec_svm_sup <- conf_matrix_sup[2,2]/sum(conf_matrix_sup[2,]))

counts <- data.frame(specificty = c(sens_svm_basic,sens_svm_ker ), sensetivity=c(spec_svm_basic,spec_svm_ker))

row.names(counts) <- c("Basic PCA","Kernelized PCA")
counts

barplot(c(counts$sensetivity, counts$specificty), main="preformance of SVM on PCA versions on RPKM data", xlab="Sensetivity and specificity", col=c("red", "blue"), legend = c("Basic PCA","Kernelized PCA", "Supervised PCA"), beside=TRUE)

```
```{r}
sample <- rownames(dat.filt.rpkm[,grepl(".*calculated",colnames(dat.filt.rpkm))])
princomp <- prcomp(dat.filt.rpkm[,grepl(".*calculated",colnames(dat.filt.rpkm))], center = T, scale = F)
pc <- data.matrix(cbind(exp_dat[sample,], princomp$x[sample,1:20 ]) )
scatter <- pc[,c("Cytogenetic_risk", "del_5", "del_7", "PC1", "PC2", "PC3")]
splom(scatter, 
      pch=16, col=1,
      pscale=0,
      xlab=NULL,
      main="Scatter plot matrix",
      diag.panel=function(x, ...){
        yLim <- current.panel.limits()$ylim
        d <- density(x, na.rm=TRUE)
        d$y <- with(d, yLim[1]+0.95*diff(yLim)*y/max(y))
        panel.lines(d, col=1)
        diag.panel.splom(x, ...)
        })



pdf(file = "del_5.pdf")
plot(princomp$x[ ,c("PC1","PC2")], bg = as.numeric(dat.filt.rpkm$del_5), pch = 21, cex = 1.5, main="del_5")
legend("topright", as.character(levels(factor(dat.filt.rpkm$del_5))), pch = 21, pt.bg =c(0,1))
dev.off()

pdf(file = "del_7.pdf")
plot(princomp$x[ ,c("PC1","PC2")], bg = as.numeric(dat.filt.rpkm$del_7), pch = 21, cex = 1.5, main="del_7")
legend("topright", as.character(levels(factor(dat.filt.rpkm$del_7))), pch = 21, pt.bg =c(0,1))
dev.off()


pdf(file = "trisomy_8.pdf")
plot(princomp$x[ ,c("PC1","PC2")], bg = as.numeric(dat.filt.rpkm$trisomy_8), pch = 21, cex = 1.5, main="trisomy_8")
legend("topright", as.character(levels(factor(dat.filt.rpkm$trisomy_8))), pch = 21, pt.bg =c(0,1))
dev.off()


pdf(file = "Molecular_risk.pdf")
plot(princomp$x[ ,c("PC1","PC2")], bg = as.numeric(dat.filt.rpkm$Molecular_risk), pch = 21, cex = 1.5, main="Molecular_risk")
legend("topright", as.character(levels(factor(dat.filt.rpkm$Molecular_risk))), pch = 21, pt.bg =c(1,2,3,4))
dev.off()



selected_gene_1 <- rownames(subset(princomp$rotation, subset=abs(princomp$rotation[,1])>abs(mean(princomp$rotation[,1])+1*sd(princomp$rotation[,1]))))
selected_gene_2 <- rownames(subset(princomp$rotation, subset=abs(princomp$rotation[,2])>abs(mean(princomp$rotation[,2])+1*sd(princomp$rotation[,2]))))
selected_gene_3 <- rownames(subset(princomp$rotation, subset=abs(princomp$rotation[,3])>abs(mean(princomp$rotation[,3])+1*sd(princomp$rotation[,3]))))
plot.new()
gene.set <- list(First_PC= selected_gene_1, Second_PC= selected_gene_2, Third_PC = selected_gene_3)
venn.plot <- venn.diagram(gene.set, filename = NULL, fill = c("red", "blue", "yellow"),force.unique)

grid.draw(venn.plot)

barplot(princomp$sd^2, main ="PCA Eigen Values")  

intersection_rpkm <- setdiff(union(union(selected_gene_1, selected_gene_2), selected_gene_3),intersect(intersect(selected_gene_1, selected_gene_2), selected_gene_3))

#comparing Read-count and RPKM data 
intersection <-  intersection_rpkm

gset <- list(RPKM= intersection_rpkm, Read_count= intersection_read)

data<- prelim.in[,grepl("_*calculated",colnames(prelim.in))]
des <- exp_dat[rownames(prelim.in),]

#stripplot(gExp ~ del_5 | intersection[1:9] , prepareData(intersection[1:9], t(data), (des)), jitter.data = TRUE, auto.key = TRUE, type = c('p', 'a'), grid = TRUE, main= "Interesting hits")
stripplot(gExp ~ Molecular_risk| intersection[1:9] , prepareData(intersection[1:4], t(data), (des)), jitter.data = TRUE, auto.key = TRUE, type = c('p', 'a'), grid = TRUE, main= "Interesting hits")


```







