Clinical Data Import and Cleaning
=================================

Read in the clinical data-sheet from the [TCGA Publication Website](https://tcga-data.nci.nih.gov/docs/publications/laml_2012/):

```{r}
library(RCurl)
library(knitr)
d <- getURL("https://tcga-data.nci.nih.gov/docs/publications/laml_2012/clinical_patient_laml.tsv")
raw_clinical_data <- read.table(text=d, header=TRUE, sep='\t')
```

This data-sheet has a lot more data than we'll actually need for this project:

```{r}
dim(raw_clinical_data)
```

Also available, as an Excel sheet, is a Supplementary table listing much of the same information, but with more annotation:

*Note that we're using the updated version as of 2013-05-13. Since the data is only available as an Excel sheet, we've done some manipulation outside of R:*

- Download the Supplementary file from [the paper website](https://tcga-data.nci.nih.gov/docs/publications/laml_2012/SuppTable01.update.2013.05.13.xlsx)
- Open in Excel, Save as CSV

```{r}
supp_d <- read.csv('SuppTable01.update.2013.05.13.csv')
dim(supp_d)
```

As a first pass, we'll try to clean the `cytogenetic_abnormality` column from the raw clinical data into a simpler categorical variable:

```{r}
summary(raw_clinical_data$cytogenetic_abnormality)
```

There are some immediate problems with this approach:

- 19 samples are missing data
- Several samples have many of the CNAs of interest
- The notation seems inconsistent between samples

*Note: Supplemenary Table 5 from the paper website might be an alternate approach here - this lists the inferred CNAs directly*

Alternate approaches possible:

1. Limit the analysis to things with a single, distinct karyotypic event (will lose lots of samples)
2. Make a 'best guess' at the most relevant alteration for each sample (messy)
3. Make several different binary variables for each of the main CNAs, e.g.:

| Patient | del(5q) | del(7q) | Trisomy 8 | t(15:17) | inv(16) | 
| ------- | ------- | ------- | --------- | -------- | ------- |
| 1 | T | F | F | F |T |
| 2 | F | F | F |T | F |

Thoughts? This would make the annotation simpler, but might complicate the analysis a bit.
